---
title: "Multiple samples"
author: "Chaoyang Ye"
date: "2024-08-28"
output: html_document
---

multiple sample stacks
correct for batch effects
use Harmony integration
```{r}
data_dir <- "data/03_session1"

dir.create(file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer"), 
           showWarnings = FALSE, recursive = TRUE)

# Spatial data adenocarcinoma prostate
download.file(url = "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0/Visium_FFPE_Human_Prostate_Cancer/Visium_FFPE_Human_Prostate_Cancer_spatial.tar.gz", 
              destfile = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer/Visium_FFPE_Human_Prostate_Cancer_spatial.tar.gz"))

# Download matrix adenocarcinoma prostate
download.file(url = "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0/Visium_FFPE_Human_Prostate_Cancer/Visium_FFPE_Human_Prostate_Cancer_raw_feature_bc_matrix.tar.gz", 
              destfile = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer/Visium_FFPE_Human_Prostate_Cancer_raw_feature_bc_matrix.tar.gz"))

dir.create(file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate"), 
           showWarnings = FALSE, recursive = TRUE)

# Spatial data normal prostate
download.file(url = "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0/Visium_FFPE_Human_Normal_Prostate/Visium_FFPE_Human_Normal_Prostate_spatial.tar.gz",
              destfile = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate/Visium_FFPE_Human_Normal_Prostate_spatial.tar.gz"))

# Download matrix normal prostate
download.file(url = "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0/Visium_FFPE_Human_Normal_Prostate/Visium_FFPE_Human_Normal_Prostate_raw_feature_bc_matrix.tar.gz",
              destfile = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate/Visium_FFPE_Human_Normal_Prostate_raw_feature_bc_matrix.tar.gz"))
```

data extraction
```{r}
# The adenocarcinoma sample
untar(tarfile = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer/Visium_FFPE_Human_Prostate_Cancer_spatial.tar.gz"), 
      exdir = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer"))

untar(tarfile = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer/Visium_FFPE_Human_Prostate_Cancer_raw_feature_bc_matrix.tar.gz"), 
      exdir = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer"))

# The normal prostate sample
untar(tarfile = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate/Visium_FFPE_Human_Normal_Prostate_spatial.tar.gz"), 
      exdir = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate"))

untar(tarfile = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate/Visium_FFPE_Human_Normal_Prostate_raw_feature_bc_matrix.tar.gz"), 
      exdir = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate"))
```

```{r}
library(Giotto)

save_dir <- "results/03_session1"
python_path <- "/usr/bin/python3"

instrs <- createGiottoInstructions(save_dir = save_dir,
                                   save_plot = TRUE,
                                   show_plot = TRUE,
                                   python_path = python_path)
```

load data
```{r}
## Healthy prostate
N_pros <- createGiottoVisiumObject(
  visium_dir = file.path(data_dir, "Visium_FFPE_Human_Normal_Prostate"),
  expr_data = "raw",
  png_name = "tissue_lowres_image.png",
  gene_column_index = 2,
  instructions = instrs
)

## Adenocarcinoma 
C_pros <- createGiottoVisiumObject(
  visium_dir = file.path(data_dir, "Visium_FFPE_Human_Prostate_Cancer"),
  expr_data = "raw",
  png_name = "tissue_lowres_image.png",
  gene_column_index = 2,
  instructions = instrs
)
```

plot
```{r}
spatPlot2D(gobject = N_pros, 
           cell_color = "in_tissue",
           show_image = TRUE, 
           point_size = 2.5,
           cell_color_code = c("black", "red"),
           point_alpha = 0.5, 
           save_param = list(save_name = "03_ses1_normal_prostate_tissue"))

spatPlot2D(gobject = C_pros, 
           cell_color = "in_tissue",
           show_image = TRUE, 
           point_size = 2.5,
           cell_color_code = c("black", "red"), 
           point_alpha = 0.5, 
           save_param = list(save_name = "03_ses1_adeno_prostate_tissue"))
```

data structure of the objects
```{r}
print(head(pDataDT(N_pros)))
print(N_pros)
```

join the two objects
```{r}
combined_pros <- joinGiottoObjects(gobject_list = list(N_pros, C_pros),
                                   gobject_names = c("NP", "CP"),
                                   join_method = "shift", x_padding = 1000)

# Printing the file structure for the individual datasets
print(head(pDataDT(combined_pros)))
print(combined_pros)

spatPlot2D(gobject = combined_pros, 
           cell_color = "in_tissue", 
           cell_color_code = c("black", "red"),
           show_image = TRUE, 
           image_name = c("NP-image", "CP-image"), 
           point_size = 1, 
           point_alpha = 0.5, 
           save_param = list(save_name = "03_ses1_combined_tissue"))
spatPlot2D(gobject = combined_pros, 
           cell_color = "in_tissue", 
           cell_color_code = c("black", "pink"),
           show_image = TRUE, 
           image_name = c("NP-image", "CP-image"),
           group_by = "list_ID", 
           point_alpha = 0.5, 
           point_size = 0.5, 
           cow_n_col = 1,
           save_param = list(save_name = "03_ses1_combined_tissue_group"))

```

reverse the process
```{r}
# Getting the cell information
combined_cells <- pDataDT(combined_pros)
np_cells <- combined_cells[list_ID == "NP"]

np_split <- subsetGiotto(combined_pros, 
                         cell_ids = np_cells$cell_ID,
                         poly_info = np_cells$cell_ID,
                         spat_unit = ":all:")

spatPlot2D(gobject = np_split, 
           cell_color = "in_tissue", 
           cell_color_code = c("black", "red"),
           show_image = TRUE, 
           point_alpha = 0.5, 
           point_size = 0.5, 
           save_param = list(save_name = "03_ses1_split_object"))
```

analyze joint objects
first filter and normalize
```{r}
# subset on in-tissue spots
metadata <- pDataDT(combined_pros)
in_tissue_barcodes <- metadata[in_tissue == 1]$cell_ID
combined_pros <- subsetGiotto(combined_pros, 
                              cell_ids = in_tissue_barcodes)

## filter
combined_pros <- filterGiotto(gobject = combined_pros,
                              expression_threshold = 1,
                              feat_det_in_min_cells = 50,
                              min_det_feats_per_cell = 500,
                              expression_values = "raw",
                              verbose = TRUE)

## normalize
combined_pros <- normalizeGiotto(gobject = combined_pros, 
                                 scalefactor = 6000)

## add gene & cell statistics
combined_pros <- addStatistics(gobject = combined_pros, 
                               expression_values = "raw")

## visualize
spatPlot2D(gobject = combined_pros, 
           cell_color = "nr_feats", 
           color_as_factor = FALSE, 
           point_size = 1, 
           show_image = TRUE, 
           image_name = c("NP-image", "CP-image"),
           save_param = list(save_name = "ses3_1_feat_expression"))
```

clustering, similar to previous exercises
```{r}
## PCA ##
combined_pros <- calculateHVF(gobject = combined_pros)

combined_pros <- runPCA(gobject = combined_pros, 
                        center = TRUE, 
                        scale_unit = TRUE)

## cluster and run UMAP ##
# sNN network (default)
combined_pros <- createNearestNetwork(gobject = combined_pros,
                                      dim_reduction_to_use = "pca", 
                                      dim_reduction_name = "pca",
                                      dimensions_to_use = 1:10, 
                                      k = 15)

# Leiden clustering
combined_pros <- doLeidenCluster(gobject = combined_pros, 
                                 resolution = 0.2,
                                 n_iterations = 200)

# UMAP
combined_pros <- runUMAP(combined_pros)
```

visualize clusters
```{r}
spatDimPlot2D(gobject = combined_pros,
              cell_color = "leiden_clus", 
              show_image = TRUE, 
              image_name = c("NP-image", "CP-image"),
              save_param = list(save_name = "ses3_1_leiden_clus"))

spatDimPlot2D(gobject = combined_pros,
              cell_color = "list_ID",
              show_image = TRUE, 
              image_name = c("NP-image", "CP-image"),
              save_param = list(save_name = "ses3_1_tissue_contribution"))
```

Harmony to integrate multiple datasets
Superior than other normalization methods
```{r}
library(harmony)

## run harmony integration
combined_pros <- runGiottoHarmony(combined_pros, 
                                  vars_use = "list_ID",
                                  do_pca = FALSE,
                                  sigma = 0.1,
                                  theta = 2,
                                  lambda = 1,
                                  nclust = NULL)
```

repeat clustering analysis
```{r}
## sNN network (default)
combined_pros <- createNearestNetwork(gobject = combined_pros,
                                      dim_reduction_to_use = "harmony",
                                      dim_reduction_name = "harmony", 
                                      name = "NN.harmony",
                                      dimensions_to_use = 1:10, 
                                      k = 15)

## Leiden clustering
combined_pros <- doLeidenCluster(gobject = combined_pros,
                                 network_name = "NN.harmony", 
                                 resolution = 0.2, 
                                 n_iterations = 1000, 
                                 name = "leiden_harmony")

# UMAP dimension reduction
combined_pros <- runUMAP(combined_pros, 
                         dim_reduction_name = "harmony", 
                         dim_reduction_to_use = "harmony", 
                         name = "umap_harmony")

spatDimPlot2D(gobject = combined_pros,
              dim_reduction_to_use = "umap", 
              dim_reduction_name = "umap_harmony",
              cell_color = "leiden_harmony", 
              show_image = TRUE, 
              image_name = c("NP-image", "CP-image"),
              spat_point_size = 1, 
              save_param = list(save_name = "leiden_clustering_harmony"))

# tissue distribution
spatDimPlot2D(gobject = combined_pros,
              dim_reduction_to_use = "umap", 
              dim_reduction_name = "umap_harmony",
              cell_color = "list_ID", 
              save_plot = TRUE,
              save_param = list(save_name = "leiden_clustering_harmony_contribution"))
```              