---
title: "Xenium practice"
author: "Chaoyang Ye"
date: "2024-08-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(Giotto)
python_path <- "/usr/bin/python3"
setwd("~/Documents/Giotto workshop")
data_path <- "data/02_session3"
save_dir <- "results/02_session3"
dir.create(save_dir, recursive = TRUE)

```

Create Giotto Xenium object
```{r}
g <- createGiottoXeniumObject(xenium_dir = data_path)
instructions(g, "save_dir") <- save_dir
instructions(g, "save_plot") <- TRUE
instructions(g, "python_path") <- python_path
```

Centroids calculation
```{r}
g <- addSpatialCentroidLocations(g, poly_info = "cell")
g <- addSpatialCentroidLocations(g, poly_info = "nucleus")
```

Visualization
```{r}
spatInSituPlotPoints(g,
    polygon_feat_type = "cell",
    feats = list(rna = head(featIDs(g))), # must be named list
    use_overlap = FALSE, 
    polygon_color = "cyan", 
    polygon_line_size = 0.1
)
```

Xenium can also be loaded stepwise
```{r}
x <- importXenium(data_path)

force(x)

x$qv <- 20 # default
tx <- x$load_transcripts()

plot(tx[[1]]$rna, dens = TRUE)
rm(tx)
```

Image metadata
```{r}
img_xml_path <- file.path(data_path, "morphology_focus", "morphology_focus_0000.xml")
omemeta <- xml2::read_xml(img_xml_path)
res <- xml2::xml_find_all(omemeta, "//d1:Channel", ns = xml2::xml_ns(omemeta))
res <- Reduce(rbind, xml2::xml_attrs(res))
rownames(res) <- NULL
res <- as.data.frame(res)
force(res)
```

Add images
```{r}
img_paths <- c(
    sprintf("data/02_session3/morphology_focus/morphology_focus_%04d.tif", 0:3),
    "data/02_session3/he_mini.tif"
)

img_list <- createGiottoLargeImageList(
    img_paths, 
    # naming is based on the channel metadata above
    names = c("DAPI", "18S", "ATP1A1/CD45/E-Cadherin", "alphaSMA/Vimentin", "HE"),
    use_rast_ext = TRUE,
    verbose = FALSE
)

# make some images brighter
img_list[[1]]@max_window <- 5000
img_list[[2]]@max_window <- 5000
img_list[[3]]@max_window <- 5000

# append images to gobject
g <- setGiotto(g, img_list)
```

Plots
```{r}
# example plots
spatInSituPlotPoints(g,
    show_image = TRUE,
    image_name = "HE",
    polygon_feat_type = "cell",
    polygon_color = "cyan",
    polygon_line_size = 0.1,
    polygon_alpha = 0
)

spatInSituPlotPoints(g,
    show_image = TRUE,
    image_name = "DAPI",
    polygon_feat_type = "nucleus",
    polygon_color = "cyan",
    polygon_line_size = 0.1,
    polygon_alpha = 0
)

spatInSituPlotPoints(g,
    show_image = TRUE,
    image_name = "18S",
    polygon_feat_type = "cell",
    polygon_color = "cyan",
    polygon_line_size = 0.1,
    polygon_alpha = 0
)

spatInSituPlotPoints(g,
    show_image = TRUE,
    image_name = "ATP1A1/CD45/E-Cadherin",
    polygon_feat_type = "nucleus",
    polygon_color = "cyan",
    polygon_line_size = 0.1,
    polygon_alpha = 0
)
```

Aggregate RNA with cell polygons
```{r}
g <- calculateOverlap(g,
    spatial_info = "cell",
    feat_info = "rna"
)

g <- overlapToMatrix(g)
```
Standard workflow
```{r}
g <- addStatistics(g, expression_values = "raw")

# very permissive filtering. Mainly for removing 0 values
g <- filterGiotto(g,
    expression_threshold = 1,
    feat_det_in_min_cells = 1,
    min_det_feats_per_cell = 5
)

g <- normalizeGiotto(g)
# overwrite original results with those for normalized values
g <- addStatistics(g)

spatInSituPlotPoints(g,
    polygon_fill = "nr_feats",
    polygon_fill_gradient_style = "sequential",
    polygon_fill_as_factor = FALSE
)

spatInSituPlotPoints(g,
    polygon_fill = "total_expr",
    polygon_fill_gradient_style = "sequential",
    polygon_fill_as_factor = FALSE
)

g <- runPCA(g, feats_to_use = NULL)
# feats_to_use = NULL since there are no HVFs calculated. Use all genes.

screePlot(g, ncp = 30)

g <- runUMAP(g, 
    dimensions_to_use = seq(15), 
    n_neighbors = 40 # default
)

plotPCA(g)
plotUMAP(g)

g <- createNearestNetwork(g,
    dimensions_to_use = seq(15), 
    k = 40
)

# takes roughly 5 min to run
g <- doLeidenCluster(g)

plotPCA_3D(g, 
    cell_color = "leiden_clus", 
    point_size = 1
)

plotUMAP(g, 
    cell_color = "leiden_clus", 
    point_size = 0.1, 
    point_shape = "no_border"
)

spatInSituPlotPoints(g,
    polygon_fill = "leiden_clus",
    polygon_fill_as_factor = TRUE,
    polygon_alpha = 0.5,
    show_image = TRUE,
    image_name = "HE"
)

```

Niche clustering
first generate spatial network
```{r}
g <- createSpatialNetwork(g,
    method = "Delaunay"
)

spatPlot2D(g, 
    point_shape = "no_border", 
    show_network = TRUE, 
    point_size = 0.1,
    point_alpha = 0.5,
    network_color = "grey"
)
```
calculate niche
```{r}
g <- calculateSpatCellMetadataProportions(g,
    spat_network = "Delaunay_network",
    metadata_column = "leiden_clus", 
    name = "leiden_niche"
)
```

k-means clustering
```{r}
# retrieve the niche info
prop_table <- getSpatialEnrichment(g, 
    name = "leiden_niche", 
    output = "data.table")

# convert to matrix
prop_matrix <- GiottoUtils::dt_to_matrix(prop_table)

# perform kmeans clustering
set.seed(1234) # make kmeans clustering reproducible

prop_kmeans <- kmeans(
    x = prop_matrix, 
    centers = 7, # controls how many clusters will be formed
    iter.max = 1000,
    nstart = 100
)

prop_kmeansDT = data.table::data.table(
    cell_ID = names(prop_kmeans$cluster),
    niche = prop_kmeans$cluster
)

# return kmeans clustering on niche to gobject
g <- addCellMetadata(g, 
    new_metadata = prop_kmeansDT, 
    by_column = TRUE,
    column_cell_ID = "cell_ID"
)

# visualize niches
spatInSituPlotPoints(g,
    show_image = TRUE,
    image_name = "HE",
    polygon_fill = "niche",
    # polygon_fill_code = getColors("Accent", 8),
    polygon_alpha = 1,
    polygon_fill_as_factor = TRUE
)

# visualize niche makeup by stacked barplot
cellmeta <- pDataDT(g)

ggplot2::ggplot(
  cellmeta, ggplot2::aes(fill = as.character(leiden_clus), 
                         y = 1, 
                         x = as.character(niche))) + 
  ggplot2::geom_bar(position = "fill", stat = "identity") + 
  ggplot2::scale_fill_manual(values = c(
    "#E7298A", "#FFED6F", "#80B1D3", "#E41A1C", "#377EB8", "#A65628", 
    "#4DAF4A", "#D9D9D9", "#FF7F00", "#BC80BD", "#666666", "#B3DE69")
  )
```

cell proximity between niches
```{r}
# uses a lot of memory
leiden_prox <- cellProximityEnrichment(g,
    cluster_column = "leiden_clus",
    spatial_network_name = "Delaunay_network",
    adjust_method = "fdr",
    number_of_simulations = 2000
)

cellProximityBarplot(g,
    CPscore = leiden_prox,
    min_orig_ints = 5, # minimum original cell-cell interactions
    min_sim_ints = 5 # minimum simulated cell-cell interactions
)
```

visualize and validate the interesting proximal niches
```{r}
# set up colors
other_cell_color <- rep("grey", 12)
int_6_8 <- int_2_9 <- other_cell_color
int_6_8[c(6, 8)] <- c("orange", "cornflowerblue") # hetero interacting niches
int_2_9[c(2, 9)] <- c("orange", "cornflowerblue")

spatInSituPlotPoints(g,
    polygon_fill = "leiden_clus",
    polygon_fill_as_factor = TRUE,
    polygon_fill_code = int_6_8,
    polygon_line_size = 0.1,
    polygon_alpha = 1,
    show_image = TRUE,
    image_name = "HE"
)

spatInSituPlotPoints(g,
    polygon_fill = "leiden_clus",
    polygon_fill_as_factor = TRUE,
    polygon_fill_code = int_2_9,
    polygon_line_size = 0.1,
    show_image = TRUE,
    polygon_alpha = 1,
    image_name = "HE"
)

```

