---
title: "Visium HD"
author: "Chaoyang Ye"
date: "2024-08-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Visium HD data
```{r}
library(Giotto)

setwd("~/Documents/Giotto workshop")

# set up paths
data_path <- "data/02_session2/"
save_dir <- "results/02_session2/"
dir.create(save_dir, recursive = TRUE)

expression_path <- file.path(data_path, 'raw_feature_bc_matrix')
expr_results <- get10Xmatrix(path_to_data = expression_path, 
                             gene_column_index = 1)

tissue_positions_path <- file.path(data_path, 'spatial/tissue_positions.parquet')
tissue_positions <- data.table::as.data.table(arrow::read_parquet(tissue_positions_path))
```

Merge expression with position data
```{r}
# convert expression matrix to minimal data.frame or data.table object
matrix_tile_dt <- data.table::as.data.table(Matrix::summary(expr_results))
genes   <- expr_results@Dimnames[[1]] # gene names
samples <- expr_results@Dimnames[[2]] # pixel name
matrix_tile_dt[, gene := genes[i]] # adding gene names as an extra column
matrix_tile_dt[, pixel := samples[j]] # adding pixel name as another extra column

# merge data.table matrix and spatial coordinates to create input for Giotto Polygons
expr_pos_data <- data.table::merge.data.table(matrix_tile_dt, 
                                             tissue_positions, 
                                             by.x = 'pixel', 
                                             by.y = 'barcode')

expr_pos_data <- expr_pos_data[,.(pixel, pxl_row_in_fullres, pxl_col_in_fullres, gene, x)]
colnames(expr_pos_data) = c('pixel', 'x', 'y', 'gene', 'count')

```

Create Giotto points (hexagon)
```{r}
giotto_points = createGiottoPoints(x = expr_pos_data[,.(x, y, gene, pixel, count)])

# create giotto polygons, here we create hexagons
hexbin400 <- tessellate(extent = ext(giotto_points), 
                        shape = 'hexagon', 
                        shape_size = 400, 
                        name = 'hex400') 
plot(hexbin400)
```

Create Giotto object by combining Giotto points and polygons
```{r}
instrs = createGiottoInstructions(
  save_dir = save_dir,
  save_plot = TRUE,
  show_plot = FALSE,
  return_plot = FALSE
)

# gpoints provides spatial gene expression information
# gpolygons provides spatial unit information (here = hexagon tiles)
visiumHD = createGiottoObjectSubcellular(gpoints = list('rna' = giotto_points),
                                         gpolygons = list('hex400' = hexbin400),
                                         instructions = instrs)

# create spatial centroids for each spatial unit (hexagon)
visiumHD = addSpatialCentroidLocations(gobject = visiumHD,
                                       poly_info = 'hex400')
```

Visualize the points in polygons
```{r}
feature_data = fDataDT(visiumHD)

spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = list('rna' = feature_data$feat_ID[1:10]), # 10 genes shown
                     show_legend = T,
                     spat_unit = 'hex400',
                     point_size = 0.25,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex400',
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.1,
                     expand_counts = TRUE, # multiple genes in the same polygon
                     count_info_column = 'count',
                     jitter = c(25,25))

spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = list('rna' = feature_data$feat_ID[10:20]),
                     show_legend = T,
                     spat_unit = 'hex400',
                     point_size = 0.25,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex400',
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.1,
                     expand_counts = TRUE, 
                     count_info_column = 'count',
                     jitter = c(25,25), 
                     plot_method = 'scattermore') # lower res, rasterized image and smaller file 
```

Bin gene expression into polygons for spatial analysis pipeline
```{r}
# calculate overlap between points and polygons
visiumHD = calculateOverlap(visiumHD,
                            spatial_info = 'hex400',
                            feat_info = 'rna')

showGiottoSpatialInfo(visiumHD)

# convert overlap results to bin by gene matrix
visiumHD = overlapToMatrix(visiumHD,
                           poly_info = 'hex400',
                           feat_info = 'rna',
                           name = 'raw')

# this action will automatically create an active spatial unit, ie. hexbin 400
activeSpatUnit(visiumHD)
```

Standard process
```{r}
# filter on gene expression matrix
visiumHD <- filterGiotto(visiumHD,
                         expression_threshold = 1,
                         feat_det_in_min_cells = 5,
                         min_det_feats_per_cell = 25)

# normalize and scale gene expression data
visiumHD <- normalizeGiotto(visiumHD, 
                            scalefactor = 1000, 
                            verbose = T)

# add cell and gene statistics
visiumHD <- addStatistics(visiumHD)
```

Visualization
```{r}
# each dot here represents a 200x200 aggregation of spatial barcodes (bin size 200)
spatPlot2D(gobject = visiumHD,
           cell_color = "nr_feats",
           color_as_factor = F,
           point_size = 2.5)

# using the polygon tiles
spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = NULL,
                     show_legend = F,
                     spat_unit = 'hex400',
                     point_size = 0.1,
                     show_polygon = TRUE,
                     use_overlap = TRUE,
                     polygon_feat_type = 'hex400', 
                     polygon_fill = 'nr_feats', 
                     polygon_fill_as_factor = F,
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.1)
```

Standard dimension reduction and clustering
```{r}
visiumHD <- calculateHVF(visiumHD, 
                         zscore_threshold = 1)
visiumHD <- runPCA(visiumHD, 
                   expression_values = 'normalized', 
                   feats_to_use = 'hvf')
screePlot(visiumHD, ncp = 30)

visiumHD <- runUMAP(visiumHD, 
                    dimensions_to_use = 1:14, 
                    n_threads = 10)

plotUMAP(gobject = visiumHD,
         point_size = 1)

# sNN network (default)
visiumHD <- createNearestNetwork(visiumHD, 
                                 dimensions_to_use = 1:14, 
                                 k = 5)

## leiden clustering ####
visiumHD <- doLeidenClusterIgraph(visiumHD, resolution = 0.5, n_iterations = 1000, spat_unit = 'hex400')

plotUMAP(gobject = visiumHD,
         cell_color = 'leiden_clus', 
         point_size = 1.5,
         show_NN_network = F, 
         edge_alpha = 0.05) # umap color based on leiden clustering

# in situ color based on leiden clustering
spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = NULL,
                     show_legend = F,
                     spat_unit = 'hex400',
                     point_size = 0.25,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex400',
                     polygon_fill_as_factor = TRUE,
                     polygon_fill = 'leiden_clus',
                     polygon_color = 'black',
                     polygon_line_size = 0.3)
```

Higher resolution analysis with hex100
```{r}
hexbin100 <- tessellate(extent = ext(visiumHD), 
                        shape = 'hexagon', 
                        shape_size = 100, 
                        name = 'hex100') 

visiumHD = setPolygonInfo(gobject = visiumHD,
                          x = hexbin100,
                          name = 'hex100',
                          initialize = T)

visiumHD = addSpatialCentroidLocations(gobject = visiumHD,
                                       poly_info = 'hex100')

activeSpatUnit(visiumHD) <- 'hex100'

spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = list('rna' = feature_data$feat_ID[1:20]),
                     show_legend = T,
                     spat_unit = 'hex100',
                     point_size = 0.1,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex100',
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.2,
                     expand_counts = TRUE, 
                     count_info_column = 'count',
                     jitter = c(25,25))

```

Combining gene expression with polygon info
```{r}
visiumHD = calculateOverlap(visiumHD,
                            spatial_info = 'hex100',
                            feat_info = 'rna')

visiumHD = overlapToMatrix(visiumHD,
                           poly_info = 'hex100',
                           feat_info = 'rna',
                           name = 'raw')

visiumHD <- filterGiotto(visiumHD,
                         expression_threshold = 1,
                         feat_det_in_min_cells = 10,
                         min_det_feats_per_cell = 10)

visiumHD <- normalizeGiotto(visiumHD, scalefactor = 1000, verbose = T)

visiumHD <- addStatistics(visiumHD)

pDataDT(visiumHD, spat_unit = 'hex100') # 6182 tiles
pDataDT(visiumHD, spat_unit = 'hex400') # 367 tiles
```

Standard processing
```{r}
## dimension reduction ####
# --------------------------- #
visiumHD <- calculateHVF(visiumHD, zscore_threshold = 1)

visiumHD <- runPCA(visiumHD, expression_values = 'normalized', feats_to_use = 'hvf')

plotPCA(visiumHD)

visiumHD <- runUMAP(visiumHD, dimensions_to_use = 1:14, n_threads = 10)

# plot UMAP, coloring cells/points based on nr_feats
plotUMAP(gobject = visiumHD,
         point_size = 2)

# sNN network (default)
visiumHD <- createNearestNetwork(visiumHD, 
                                 dimensions_to_use = 1:14, 
                                 k = 5)

## leiden clustering ####
visiumHD <- doLeidenClusterIgraph(visiumHD, resolution = 0.2, n_iterations = 1000)

plotUMAP(gobject = visiumHD,
         cell_color = 'leiden_clus', 
         point_size = 1.5,
         show_NN_network = F, 
         edge_alpha = 0.05)

spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = NULL,
                     show_legend = F,
                     spat_unit = 'hex100',
                     point_size = 0.5,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex100',
                     polygon_fill_as_factor = TRUE,
                     polygon_fill = 'leiden_clus',
                     polygon_color = 'black',
                     polygon_line_size = 0.3)

```

Spatial expression pattern
```{r}
featData = fDataDT(visiumHD)
hvf_genes = featData[hvf == 'yes']$feat_ID

visiumHD = createSpatialNetwork(visiumHD,
                                name = 'kNN_network',
                                spat_unit = 'hex100', 
                                method = 'kNN',
                                k = 8)

ranktest = binSpect(visiumHD,
                    spat_unit = 'hex100',
                    subset_feats = hvf_genes,
                    bin_method = 'rank',
                    calc_hub = FALSE,
                    do_fisher_test = TRUE,
                    spatial_network_name = 'kNN_network')

set0 = ranktest[high_expr < 50][1:2]$feats
set1 = ranktest[high_expr > 50 & high_expr < 100][1:2]$feats
set2 = ranktest[high_expr > 100 & high_expr < 200][1:2]$feats
set3 = ranktest[high_expr > 200 & high_expr < 400][1:2]$feats
set4 = ranktest[high_expr > 400 & high_expr < 1000][1:2]$feats
set5 = ranktest[high_expr > 1000][1:2]$feats

spatFeatPlot2D(visiumHD,
               expression_values = 'scaled',
               feats = c(set0, set1, set2),
               gradient_style = "sequential",
               cell_color_gradient = c('blue', 'white', 'yellow', 'orange', 'red', 'darkred'),
               cow_n_col = 2, point_size = 1)

spatFeatPlot2D(visiumHD,
               expression_values = 'scaled',
               feats = c(set3, set4, set5),
               gradient_style = "sequential",
               cell_color_gradient = c('blue', 'white', 'yellow', 'orange', 'red', 'darkred'),
               cow_n_col = 2, point_size = 1)
```

Spatial co-expression modules
```{r}
ext_spatial_genes = ranktest[adj.p.value < 0.001]$feats

spat_cor_netw_DT = detectSpatialCorFeats(visiumHD,
                                         method = 'network',
                                         spatial_network_name = 'kNN_network',
                                         subset_feats = ext_spatial_genes)

# cluster spatial genes
spat_cor_netw_DT = clusterSpatialCorFeats(spat_cor_netw_DT, 
                                          name = 'spat_netw_clus', 
                                          k = 16)

# visualize clusters
heatmSpatialCorFeats(visiumHD,
                     spatCorObject = spat_cor_netw_DT,
                     use_clus_name = 'spat_netw_clus',
                     heatmap_legend_param = list(title = NULL))
```

Metagene enrichment scores for clusters
```{r}
# create metagene enrichment score for clusters
cluster_genes_DT = showSpatialCorFeats(spat_cor_netw_DT, 
                                       use_clus_name = 'spat_netw_clus', 
                                       show_top_feats = 1)
cluster_genes = cluster_genes_DT$clus; names(cluster_genes) = cluster_genes_DT$feat_ID

visiumHD = createMetafeats(visiumHD,
                           expression_values = 'normalized',
                           feat_clusters = cluster_genes,
                           name = 'cluster_metagene')

showGiottoSpatEnrichments(visiumHD)

spatCellPlot(visiumHD,
             spat_enr_names = 'cluster_metagene',
             gradient_style = "sequential",
             cell_color_gradient = c('blue', 'white', 'yellow', 'orange', 'red', 'darkred'),
             cell_annotation_values = as.character(c(1:4)),
             point_size = 1, cow_n_col = 2)
```

Plot spatial gene groups
```{r}
balanced_genes = getBalancedSpatCoexpressionFeats(spatCorObject = spat_cor_netw_DT, 
                                                  maximum = 5)
selected_feats = names(balanced_genes)

# give genes from same cluster same color
distinct_colors = getDistinctColors(n = 20)
names(distinct_colors) = 1:20

my_colors = distinct_colors[balanced_genes]
names(my_colors) = names(balanced_genes)

spatInSituPlotPoints(visiumHD,
                     show_image = F,
                     feats = list('rna' = selected_feats), 
                     feats_color_code = my_colors,
                     show_legend = F,
                     spat_unit = 'hex100',
                     point_size = 0.20,
                     show_polygon = FALSE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex100',
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.01,
                     expand_counts = TRUE, 
                     count_info_column = 'count',
                     jitter = c(25,25))
```

Even finer resolution with hex25 for a subsection of the image
```{r}
# select subsection
visiumHD_subset = subsetGiottoLocs(gobject = visiumHD,
                                   x_min = 16000, 
                                   x_max = 20000, 
                                   y_min = 44250,
                                   y_max = 45500)
spatInSituPlotPoints(visiumHD_subset,
                     show_image = F,
                     feats = NULL,
                     show_legend = F,
                     spat_unit = 'hex100',
                     point_size = 0.5,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex100',
                     polygon_fill_as_factor = TRUE,
                     polygon_fill = 'leiden_clus',
                     polygon_color = 'black',
                     polygon_line_size = 0.3)
spatInSituPlotPoints(visiumHD_subset,
                     show_image = F,
                     feats = list('rna' = selected_feats), 
                     feats_color_code = my_colors,
                     show_legend = F,
                     spat_unit = 'hex100',
                     point_size = 0.40,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex100',
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.05,
                     jitter = c(25,25))
```

Create hex25 tessellation
```{r}
hexbin25 <- tessellate(extent = ext(visiumHD_subset@feat_info$rna), 
                       shape = 'hexagon', 
                       shape_size = 25, 
                       name = 'hex25') 

visiumHD_subset = setPolygonInfo(gobject = visiumHD_subset,
                                 x = hexbin25,
                                 name = 'hex25',
                                 initialize = T)

showGiottoSpatialInfo(visiumHD_subset)

visiumHD_subset = addSpatialCentroidLocations(gobject = visiumHD_subset,
                                              poly_info = 'hex25')

activeSpatUnit(visiumHD_subset) <- 'hex25'

spatInSituPlotPoints(visiumHD_subset,
                     show_image = F,
                     feats = list('rna' = selected_feats), 
                     feats_color_code = my_colors,
                     show_legend = F,
                     spat_unit = 'hex25',
                     point_size = 0.40,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex25',
                     polygon_bg_color = NA,
                     polygon_color = 'white',
                     polygon_line_size = 0.05,
                     jitter = c(25,25))
```

Standard preprocessing, dimension reduction and clustering
```{r}
visiumHD_subset = calculateOverlap(visiumHD_subset,
                                   spatial_info = 'hex25',
                                   feat_info = 'rna')

showGiottoSpatialInfo(visiumHD_subset)


# convert overlap results to bin by gene matrix
visiumHD_subset = overlapToMatrix(visiumHD_subset,
                                  poly_info = 'hex25',
                                  feat_info = 'rna',
                                  name = 'raw')

visiumHD_subset <- filterGiotto(visiumHD_subset,
                                expression_threshold = 1,
                                feat_det_in_min_cells = 3,
                                min_det_feats_per_cell = 5)

activeSpatUnit(visiumHD_subset)


# normalize
visiumHD_subset <- normalizeGiotto(visiumHD_subset, scalefactor = 1000, verbose = T)

# add statistics
visiumHD_subset <- addStatistics(visiumHD_subset)

feature_data = fDataDT(visiumHD_subset)

visiumHD_subset <- calculateHVF(visiumHD_subset, zscore_threshold = 1)

n_25_percent <- round(length(spatIDs(visiumHD_subset, 'hex25')) * 0.25)

# pca projection on subset
visiumHD_subset <- runPCAprojection(
  gobject = visiumHD_subset,
  spat_unit = "hex25",
  feats_to_use = 'hvf',
  name = 'pca.projection',
  set_seed = TRUE,
  seed_number = 12345,
  random_subset = n_25_percent
)

showGiottoDimRed(visiumHD_subset)
plotPCA(visiumHD_subset, dim_reduction_name = 'pca.projection')

# umap projection on subset
visiumHD_subset <- runUMAPprojection(
  gobject = visiumHD_subset,
  spat_unit = "hex25",
  dim_reduction_to_use = 'pca',
  dim_reduction_name = "pca.projection",
  dimensions_to_use = 1:10,
  name = "umap.projection",
  random_subset = n_25_percent, 
  n_neighbors = 10,
  min_dist = 0.005,
  n_threads = 4
)

showGiottoDimRed(visiumHD_subset)

# plot UMAP, coloring cells/points based on nr_feats
plotUMAP(gobject = visiumHD_subset,
         point_size = 1, 
         dim_reduction_name = 'umap.projection')

# subset to smaller giotto object
set.seed(1234)
subset_IDs = sample(x = spatIDs(visiumHD_subset, 'hex25'), size = n_25_percent)

temp_gobject = subsetGiotto(
  gobject = visiumHD_subset,
  spat_unit = 'hex25',
  cell_ids = subset_IDs
)


# hierarchical clustering
temp_gobject = doHclust(gobject = temp_gobject, 
                        spat_unit = 'hex25', 
                        k = 8, name = 'sub_hclust', 
                        dim_reduction_to_use = 'pca', 
                        dim_reduction_name = 'pca.projection', 
                        dimensions_to_use = 1:10)


# show umap
dimPlot2D(
  gobject = temp_gobject,
  point_size = 2.5,
  spat_unit = 'hex25',
  dim_reduction_to_use = 'umap',
  dim_reduction_name = 'umap.projection',
  cell_color = 'sub_hclust'
)
```

Project to full dataset
```{r}
# project clusterings back to full dataset
visiumHD_subset <- doClusterProjection(
  target_gobject = visiumHD_subset,
  source_gobject = temp_gobject,
  spat_unit = "hex25",
  source_cluster_labels = "sub_hclust",
  reduction_method = 'pca',
  reduction_name = 'pca.projection',
  prob = FALSE,
  knn_k = 5,
  dimensions_to_use = 1:10
)

pDataDT(visiumHD_subset)

dimPlot2D(
  gobject = visiumHD_subset,
  point_size = 1.5,
  spat_unit = 'hex25',
  dim_reduction_to_use = 'umap',
  dim_reduction_name = 'umap.projection',
  cell_color = 'knn_labels'
)
```

Plot hex25 cluster in the subset
```{r}
spatInSituPlotPoints(visiumHD_subset,
                     show_image = F,
                     feats = NULL,
                     show_legend = F,
                     spat_unit = 'hex25',
                     point_size = 0.5,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex25',
                     polygon_fill_as_factor = TRUE,
                     polygon_fill = 'knn_labels',
                     polygon_color = 'black',
                     polygon_line_size = 0.3)
```

Niche clustering: cell and its surrounding cells forming a micro environment
```{r}
visiumHD_subset = createSpatialNetwork(visiumHD_subset,
                                name = 'kNN_network',
                                spat_unit = 'hex25', 
                                method = 'kNN',
                                k = 6)

pDataDT(visiumHD_subset)
visiumHD_subset = calculateSpatCellMetadataProportions(gobject = visiumHD_subset,
                                                       spat_unit = 'hex25',
                                                       feat_type = 'rna',
                                                       metadata_column = 'knn_labels',
                                                       spat_network = 'kNN_network')

prop_table = getSpatialEnrichment(visiumHD_subset, name = 'proportion', output = 'data.table')
prop_matrix = GiottoUtils:::dt_to_matrix(prop_table)

set.seed(1234)
prop_kmeans = kmeans(x = prop_matrix, centers = 10, iter.max = 1000, nstart = 100)
prop_kmeansDT = data.table::data.table(cell_ID = names(prop_kmeans$cluster), niche = prop_kmeans$cluster)

visiumHD_subset = addCellMetadata(visiumHD_subset, 
                                  new_metadata = prop_kmeansDT, 
                                  by_column = T, 
                                  column_cell_ID = 'cell_ID')
pDataDT(visiumHD_subset)

spatInSituPlotPoints(visiumHD_subset,
                     show_image = F,
                     feats = NULL,
                     show_legend = F,
                     spat_unit = 'hex25',
                     point_size = 0.5,
                     show_polygon = TRUE,
                     use_overlap = FALSE,
                     polygon_feat_type = 'hex25',
                     polygon_fill_as_factor = TRUE,
                     polygon_fill = 'niche',
                     polygon_color = 'black',
                     polygon_line_size = 0.3)
```


